---
title: "Prediction and Model Selection"
---


## Learning outcomes {.unnumbered}

By the end of this chapter the reader should be able to:

- Distinguish *causal inference* from *prediction* as two different econometric goals, and explain why a model that predicts well is not automatically a good causal model.
- Compute the point prediction $\hat y_0$ for a new observation with regressor values $x_0$, both in the simple and multiple linear regression settings.
- Construct and interpret the two distinct interval estimates: the **confidence interval** for the conditional mean $\mathbb{E}[y \mid x_0]$, and the **prediction interval** for an individual outcome $y_0$.
- Explain why the prediction interval is always wider than the confidence interval for the mean, and identify the conditions under which both intervals are wider.
- Compare nested models using $R^2$, adjusted $R^2$, the Akaike (AIC) and Bayesian (BIC) information criteria, and the nested $F$-test.
- Use `predict()`, `AIC()`, `BIC()`, and `anova()` in R to predict from a fitted model and to choose among competing specifications.

## Motivating empirical question {.unnumbered}

> *Two recruiters interview the same candidate --- a 24-year-old with 16 years of education, 5 years of experience, and 2 years of tenure --- and both fit the same wage regression to the same data. The first reports a 95% interval of [\$5.7, \$7.2] for the expected wage; the second reports [\$0.5, \$12.4]. Both are correct. Which interval is which?*

The answer turns on a distinction that is easy to miss: the first recruiter is reporting uncertainty about the *average* wage of people like this candidate, while the second is reporting uncertainty about *this* candidate's actual wage. The same fitted model produces both numbers --- they answer different questions. This chapter is about that distinction and the related question of how to choose between competing models in the first place.

## 5.1 Causality vs. prediction

Up to Chapter 4 we have used regression for one purpose: to estimate the *ceteris paribus* effect of $x$ on $y$. The coefficient $\beta_j$ was the object of interest, and the entire apparatus of Gauss--Markov assumptions, $t$-tests, and $F$-tests was in the service of getting $\hat\beta_j$ right --- unbiased, efficient, and with valid standard errors.

Econometrics also has a second, equally legitimate use: **prediction**. Here we do not particularly care about the value of any individual $\beta_j$; we care about the accuracy of the fitted value $\hat y_0$ for a new observation.

The two goals overlap but are not the same:

- **Causal inference.** The target is $\partial \mathbb{E}[y \mid x] / \partial x_j$. The decisive assumption is MLR.4 (zero conditional mean): $\mathbb{E}[u \mid x_1, \dots, x_k] = 0$. Coefficients are interpreted; the typical statement is "a one-unit increase in $x_j$ *causes* an expected change of $\beta_j$ in $y$, holding other factors constant".
- **Prediction.** The target is $\hat y_0$ for a new $x_0$. The decisive assumption is *stability*: the new $(y_0, x_0)$ is drawn from the same population as the sample. Coefficients need not be interpretable; we want regressors with strong predictive power and high (in-sample and out-of-sample) fit.

A model can be excellent for one purpose and poor for the other. A predictor of road accidents that includes "ice-cream sales" as a regressor may work well because ice cream proxies for hot weather; it is useless for causal policy advice about ice cream. Conversely, an instrumental-variables estimate of the return to schooling might identify a clean causal parameter from a tiny slice of variation in the data, while leaving most of the variability in wages unexplained.

::: {.callout-warning}
## Common mistake: choosing the model with the highest $R^2$ and reading off causal effects

A high $R^2$ tells you that the fitted values track $y$ well. It does **not** tell you that any individual coefficient measures a causal effect. The identification strategy --- random assignment, an instrument, a natural experiment, or a credible argument that $\mathbb{E}[u \mid x] = 0$ --- is what licences causal interpretation. Fit-based selection criteria such as adjusted $R^2$, AIC, and BIC are tools for prediction; they cannot rescue a model that suffers from omitted-variable bias.
:::

## 5.2 Point prediction

Consider, for clarity, the simple linear regression (SLR) model:

$$
y_i = \beta_0 + \beta_1 x_i + u_i, \qquad \mathbb{E}[u_i \mid x_i] = 0.
$$

Given the OLS estimates $(\hat\beta_0, \hat\beta_1)$ and a new value $x_0$ of the regressor, the **point prediction** is just the fitted value of the regression at $x_0$:

$$
\hat y_0 = \hat\beta_0 + \hat\beta_1 x_0.
$$

The same construction extends, term by term, to multiple linear regression (MLR). If $x_0 = (1, x_{01}, x_{02}, \dots, x_{0k})$ is the row vector of regressor values for the new observation, then

$$
\hat y_0 = \hat\beta_0 + \hat\beta_1 x_{01} + \hat\beta_2 x_{02} + \cdots + \hat\beta_k x_{0k}.
$$

How good is this prediction? The natural quantity is the **forecast error**:

$$
f \equiv y_0 - \hat y_0 = (\beta_0 + \beta_1 x_0 + u_0) - (\hat\beta_0 + \hat\beta_1 x_0).
$$

Taking the conditional expectation under MLR.1--MLR.4,

$$
\mathbb{E}[f \mid x_0] = (\beta_0 + \beta_1 x_0) - \mathbb{E}[\hat\beta_0 + \hat\beta_1 x_0] = 0.
$$

So $\hat y_0$ is an **unbiased predictor** of $y_0$. In fact, under SLR.1--SLR.5 (and their MLR counterparts), one can show that $\hat y_0$ has the smallest variance among all linear unbiased predictors: it is the **Best Linear Unbiased Predictor (BLUP)**.

::: {.callout-warning}
## $\hat y_0$ plays a dual role

The same number $\hat y_0$ is *both* the point prediction of the individual outcome $y_0$ *and* the estimate of the conditional mean $\mathbb{E}[y \mid x_0]$. The two interpretations of $\hat y_0$ are numerically identical, but they have different *uncertainty*. The next section makes this concrete.
:::

## 5.3 Forecast variance and prediction intervals

A point estimate is rarely enough. We need an interval that quantifies uncertainty --- and there are two distinct uncertainties to quantify.

### 5.3.1 Confidence interval for the conditional mean $\mathbb{E}[y \mid x_0]$

The conditional mean $\mathbb{E}[y \mid x_0] = \beta_0 + \beta_1 x_0$ is a fixed (non-random) population quantity. Its estimator is $\hat y_0 = \hat\beta_0 + \hat\beta_1 x_0$, which is random because the OLS estimates are random. Its variance is

$$
\operatorname{Var}(\hat y_0 \mid x) = \sigma^2 \left[\frac{1}{n} + \frac{(x_0 - \bar x)^2}{\mathrm{SST}_x}\right],
$$

where $\mathrm{SST}_x = \sum_{i=1}^{n} (x_i - \bar x)^2$ is the total sum of squares of the regressor. Replacing $\sigma^2$ by the estimator $\hat\sigma^2 = \mathrm{SSR}/(n - k - 1)$, the $100(1 - \alpha)\%$ confidence interval for the *mean* prediction is

$$
\hat y_0 \;\pm\; t_c \cdot \hat\sigma \sqrt{\frac{1}{n} + \frac{(x_0 - \bar x)^2}{\mathrm{SST}_x}},
$$

with $t_c$ the appropriate critical value of a $t$-distribution with $n - k - 1$ degrees of freedom.

### 5.3.2 Prediction interval for an individual outcome $y_0$

The new outcome $y_0 = \beta_0 + \beta_1 x_0 + u_0$ is random because *both* the OLS estimates *and* the new error $u_0$ are random. Assuming $u_0$ is independent of the sample,

$$
\operatorname{Var}(y_0 - \hat y_0 \mid x) = \sigma^2 \left[1 + \frac{1}{n} + \frac{(x_0 - \bar x)^2}{\mathrm{SST}_x}\right].
$$

The extra "$1$" inside the brackets is the contribution of $u_0$ --- the irreducible part of the new draw that no amount of data could ever predict. The $100(1 - \alpha)\%$ **prediction interval** for $y_0$ is

$$
\hat y_0 \;\pm\; t_c \cdot \hat\sigma \sqrt{1 + \frac{1}{n} + \frac{(x_0 - \bar x)^2}{\mathrm{SST}_x}}.
$$

In MLR the same logic gives $\operatorname{Var}(y_0 - \hat y_0 \mid X) = \sigma^2 [1 + x_0'(X'X)^{-1} x_0]$, with the CI for the mean obtained by dropping the "$1$" in the brackets.

::: {.callout-note}
## The two intervals at a glance

| Quantity targeted | Variance ingredient | Interval | Width |
|-------------------|---------------------|----------|-------|
| Mean $\mathbb{E}[y \mid x_0]$ | $1/n + (x_0 - \bar x)^2 / \mathrm{SST}_x$ | Confidence interval | Narrower |
| Individual $y_0$ | $1 + 1/n + (x_0 - \bar x)^2 / \mathrm{SST}_x$ | Prediction interval | Wider |

The PI is always wider than the CI. The gap is governed by the "1": the irreducible variance of the new draw $u_0$.
:::

A few more observations on the formulas:

- Both variances shrink as $n$ grows --- with infinite data the CI for the mean would collapse to a point, but the PI would not, because the "1" never goes away.
- Both variances grow as $x_0$ moves *away from the sample mean* $\bar x$. Predictions are most precise near the centre of the data; predicting at the edges (or extrapolating beyond them) is dangerous.
- Both variances shrink as $\mathrm{SST}_x$ grows. More spread-out regressors give more leverage to pin down the slope, which in turn tightens prediction.

### 5.3.3 Worked numerical example

Suppose we have estimated the SLR

$$
\hat y = 2.0 + 0.7\,x,
$$

with sample size $n = 100$, $\bar x = 12$, $\mathrm{SST}_x = 200$, and $\hat\sigma^2 = 4$ (so $\hat\sigma = 2$). We want to predict at $x_0 = 16$. Using $t_c \approx 1.98$ for a 95% interval with 98 degrees of freedom:

- **Point prediction.** $\hat y_0 = 2.0 + 0.7 \times 16 = 13.2$.

- **CI for the mean** $\mathbb{E}[y \mid x_0]$.
$$
\mathrm{se}(\hat y_0) = \hat\sigma \sqrt{\tfrac{1}{100} + \tfrac{(16-12)^2}{200}} = 2\sqrt{0.01 + 0.08} = 0.6.
$$
$$
\text{95\% CI:} \quad 13.2 \pm 1.98 \times 0.6 = [12.01,\ 14.39].
$$

- **Prediction interval** for $y_0$.
$$
\mathrm{se}(f) = \hat\sigma \sqrt{1 + \tfrac{1}{100} + \tfrac{(16-12)^2}{200}} = 2\sqrt{1.09} \approx 2.088.
$$
$$
\text{95\% PI:} \quad 13.2 \pm 1.98 \times 2.088 = [9.06,\ 17.34].
$$

The CI is roughly $\pm 1.20$ wide around the point prediction; the PI is roughly $\pm 4.14$ wide. The ratio is set by the relative size of the "$1$" against $1/n + (x_0 - \bar x)^2/\mathrm{SST}_x = 0.09$. Here the new-draw uncertainty dwarfs the parameter uncertainty by an order of magnitude.

## 5.4 Prediction in MLR

Real wage models include more than one regressor. Nothing essential changes in the construction of $\hat y_0$ and the two intervals --- the formulas just become vector--matrix versions of the SLR formulas.

For a row vector $x_0 = (1, x_{01}, \dots, x_{0k})$,

$$
\hat y_0 = x_0' \hat\beta, \qquad \operatorname{Var}(y_0 - \hat y_0 \mid X) = \sigma^2 \left[1 + x_0' (X'X)^{-1} x_0 \right].
$$

Dropping the "$1$" gives the variance of $\hat y_0$ as an estimator of $\mathbb{E}[y \mid x_0]$. In R, the function `predict()` does all of this automatically with `interval = "confidence"` or `interval = "prediction"`. The user supplies the new $x_0$ as a data frame; the function returns the fit and the bounds. We try this on `wage1` in the lab.

::: {.callout-warning}
## Don't extrapolate

The prediction formulas assume that $(y_0, x_0)$ comes from the *same population* as the sample. If $x_0$ is far outside the range of the regressors actually observed --- predicting wages for someone with 30 years of education and 80 years of experience, say --- both intervals can be technically computed, but neither is trustworthy. There is no information in the sample about what happens at those values. Stay inside the support of the data.
:::

## 5.5 Model selection: $R^2$, adjusted $R^2$, AIC, BIC

Suppose we have several candidate regression models for the same outcome. How do we choose? Four criteria are commonly used.

### 5.5.1 $R^2$ and the garbage-variable problem

The coefficient of determination

$$
R^2 = \frac{\mathrm{SSE}}{\mathrm{SST}} = 1 - \frac{\mathrm{SSR}}{\mathrm{SST}}
$$

measures the proportion of the variation of $y$ explained by the model. It has one fatal flaw as a model-selection device: **adding any regressor cannot reduce $R^2$.** Even a pure noise regressor will, almost surely, slightly increase $R^2$, because OLS can always find a tiny linear combination of the noise that fits some residual variation by chance.

If we picked our model by maximising $R^2$, we would include every regressor we could lay our hands on --- including dozens of irrelevant ones. The model would fit the sample beautifully and predict horribly out of sample. This is the classic *overfitting* problem. The slides illustrate it by simulating a model with two genuine regressors $X_1, X_2$ and then adding 196 "garbage" variables of pure noise: $R^2$ climbs steadily towards 1 as garbage piles up, even though none of the new variables has any predictive content.

### 5.5.2 Adjusted $R^2$

A simple fix is to penalise complexity. The **adjusted $R^2$** is

$$
\bar R^2 = 1 - \frac{\mathrm{SSR} / (n - k - 1)}{\mathrm{SST} / (n - 1)},
$$

where $k$ is the number of regressors (excluding the intercept). The numerator $\mathrm{SSR}/(n - k - 1)$ is $\hat\sigma^2$, the unbiased estimator of the error variance. Adding a regressor reduces $\mathrm{SSR}$ but also reduces $n - k - 1$; $\bar R^2$ goes up only if the gain in fit beats the cost of the extra parameter.

A few useful facts:

- $\bar R^2 \le R^2$ always, with equality only when $k = 0$.
- Unlike $R^2$, $\bar R^2$ can *decrease* when an irrelevant regressor is added --- and in the garbage simulation, it does.
- $\bar R^2$ can in principle be negative if the model is so poor that $\hat\sigma^2$ exceeds the sample variance of $y$.

### 5.5.3 AIC and BIC

The **Akaike Information Criterion** and the **Bayesian Information Criterion** combine model fit (a log-likelihood) with a complexity penalty:

$$
\mathrm{AIC} = -2 \ln \hat L + 2(k + 1), \qquad \mathrm{BIC} = -2 \ln \hat L + (k + 1) \ln n.
$$

For a linear model with normally distributed errors the log-likelihood is a monotone transformation of $-\mathrm{SSR}/n$, so up to additive constants

$$
\mathrm{AIC} \approx n \ln\!\left(\frac{\mathrm{SSR}}{n}\right) + 2(k+1), \qquad \mathrm{BIC} \approx n \ln\!\left(\frac{\mathrm{SSR}}{n}\right) + (k+1) \ln n.
$$

**Lower AIC and BIC indicate better models.** Both criteria balance fit (the SSR term) against complexity (the second term). The only difference is the size of the penalty: AIC charges $2$ per extra parameter, BIC charges $\ln n$. For $n \ge 8$, $\ln n > 2$, so BIC penalises complexity more strongly and tends to select more parsimonious models. AIC is built to minimise prediction error; BIC is built to identify the "true" model when one exists.

::: {.callout-warning}
## AIC and BIC require the same dataset and the same dependent variable

If two models use different dependent variables (e.g.\ `wage` and `log(wage)`), or are estimated on different sub-samples, their AIC and BIC are *not* comparable --- the log-likelihoods are on different scales. The same warning applies to $R^2$ and $\bar R^2$.
:::

### 5.5.4 Nested $F$-test

When one model is nested inside another --- the smaller model is obtained by setting some coefficients of the larger model to zero --- the **$F$-test of joint restrictions** (Chapter 4) is the formal test of "are these extra regressors jointly informative?". In R, `anova(small, big)` does it. The test answers a different question from AIC/BIC: it asks whether the data *reject* the smaller model in favour of the larger, not which model is "best" by some predictive criterion.

### 5.5.5 Putting it together

In practice we report all four criteria side by side. They usually agree on the broad ranking; when they disagree, the disagreement is informative:

- If $R^2$ rises but $\bar R^2$, AIC, and BIC all worsen, the new regressors are noise.
- If $\bar R^2$ and AIC favour the larger model but BIC favours the smaller, the extra regressors are weakly predictive; with a large sample BIC may be over-penalising. In a *purely predictive* exercise, lean towards AIC; for *parsimony* or for picking a "true" model, lean towards BIC.

And, always: model selection criteria choose the model that *fits* best on the criterion at hand. They do not choose the model that *identifies a causal parameter* best. That choice belongs to the identification strategy.

## 5.6 Lab: Prediction and Model Selection in R

We work with the `wage1` dataset of 526 U.S.\ workers (Chapter 1). The goal is to (i) fit a sequence of nested wage models, (ii) compare them using $\bar R^2$, AIC, BIC, and an $F$-test, (iii) produce point predictions and the two interval estimates for a specific worker, and (iv) visualise the difference between the confidence band for the mean and the prediction band for an individual.

```{r ch05-setup}
#| message: false
library(wooldridge)
data("wage1")
```

### Step 1 --- Fit four nested wage models

We progressively enrich the specification.

```{r ch05-models}
model1 <- lm(wage ~ educ, data = wage1)
model2 <- lm(wage ~ educ + exper, data = wage1)
model3 <- lm(wage ~ educ + exper + tenure, data = wage1)
model4 <- lm(wage ~ educ + exper + tenure + female + married, data = wage1)
```

### Step 2 --- Compare the four models

Collect $R^2$, $\bar R^2$, AIC, and BIC in a single table.

```{r ch05-criteria}
tab <- sapply(list(m1 = model1, m2 = model2, m3 = model3, m4 = model4),
              function(m) c(
                R2    = summary(m)$r.squared,
                AdjR2 = summary(m)$adj.r.squared,
                AIC   = AIC(m),
                BIC   = BIC(m)))
round(tab, 3)
```

A few things to read off this table:

- $R^2$ rises *monotonically* from `model1` to `model4`. That is mechanical: adding regressors cannot decrease $R^2$.
- $\bar R^2$ also rises across the four models, indicating that each newly added regressor contributes more fit than its complexity cost.
- AIC and BIC both *decrease* as we go from `model1` to `model4` (lower is better). All four criteria agree: of these four specifications, `model4` is preferred.

### Step 3 --- A nested $F$-test

Are `female` and `married` *jointly* informative once `educ`, `exper`, and `tenure` are in the model?

```{r ch05-anova}
anova(model3, model4)
```

The $F$-statistic compares the two SSRs; if its $p$-value is small, we reject the joint null $H_0:\beta_{\text{female}} = \beta_{\text{married}} = 0$. In `wage1` it is very small, confirming the AIC/BIC preference for `model4` with a formal test.

### Step 4 --- The garbage-variable demonstration

Now we deliberately add three pure-noise regressors to `model4` and watch what happens to each criterion.

```{r ch05-garbage}
set.seed(42)
wage1$noise1 <- rnorm(nrow(wage1))
wage1$noise2 <- rnorm(nrow(wage1))
wage1$noise3 <- rnorm(nrow(wage1))

model_garbage <- lm(wage ~ educ + exper + tenure + female + married +
                      noise1 + noise2 + noise3, data = wage1)

rbind(
  model4        = c(R2 = summary(model4)$r.squared,
                    AdjR2 = summary(model4)$adj.r.squared,
                    AIC = AIC(model4), BIC = BIC(model4)),
  model_garbage = c(R2 = summary(model_garbage)$r.squared,
                    AdjR2 = summary(model_garbage)$adj.r.squared,
                    AIC = AIC(model_garbage), BIC = BIC(model_garbage)))
```

Inspect the output: $R^2$ has risen slightly (as it must), but $\bar R^2$ has *fallen*, and both AIC and BIC have *increased*. The three pure-noise regressors look superficially helpful to plain $R^2$; every penalised criterion sees through them. This is the entire point of using $\bar R^2$, AIC, or BIC rather than raw $R^2$ to choose a model.

### Step 5 --- Point prediction for a specific worker

Take the candidate from the chapter opening: 16 years of education, 5 years of experience, 2 years of tenure, female, unmarried. We use `model4`.

```{r ch05-point}
new_worker <- data.frame(educ = 16, exper = 5, tenure = 2,
                         female = 1, married = 0)
predict(model4, newdata = new_worker)
```

This is $\hat y_0$ --- the recruiter's single-number wage prediction.

### Step 6 --- The two intervals

The same `predict()` function delivers the CI for the mean and the PI for the individual.

```{r ch05-intervals}
predict(model4, newdata = new_worker,
        interval = "confidence", level = 0.95)
predict(model4, newdata = new_worker,
        interval = "prediction", level = 0.95)
```

The first interval is the recruiter's answer to "what is the average wage of women with these characteristics?". The second is the answer to "what wage might *this* woman actually earn?". The PI is much wider --- in this dataset, an order of magnitude wider --- because it includes the irreducible variance of the new draw $u_0$.

### Step 7 --- A manual sanity check

To make the formulas concrete, compute the CI by hand and compare:

```{r ch05-manual}
yhat <- predict(model4, newdata = new_worker, se.fit = TRUE)
tc   <- qt(0.975, df = model4$df.residual)
ci_manual <- c(yhat$fit - tc * yhat$se.fit,
               yhat$fit + tc * yhat$se.fit)
ci_manual

# Now the prediction interval, using SE(f)^2 = SE(yhat)^2 + sigma_hat^2
sig2 <- summary(model4)$sigma^2
se_pred <- sqrt(yhat$se.fit^2 + sig2)
pi_manual <- c(yhat$fit - tc * se_pred,
               yhat$fit + tc * se_pred)
pi_manual
```

These should match (to rounding) the bounds returned by `predict()` in Step 6. The decomposition $\operatorname{Var}(y_0 - \hat y_0) = \operatorname{Var}(\hat y_0) + \sigma^2$ is the operational version of "two sources of uncertainty: the parameters, and the new draw".

### Step 8 --- Visualising CI vs.\ PI across `educ`

The two-interval distinction is easiest to see in a picture. We compute both bands across a grid of education values (holding the other regressors fixed) and plot them on top of the fitted line.

```{r ch05-bands}
#| fig-cap: "Fitted wage (solid blue) with the 95% confidence band for the conditional mean (dashed blue) and the 95% prediction band for individuals (dashed red), across years of education. Other regressors fixed at exper=5, tenure=3, female=0, married=1."
grid <- data.frame(educ = seq(0, 20, by = 0.5),
                   exper = 5, tenure = 3,
                   female = 0, married = 1)
ci_band <- predict(model4, newdata = grid, interval = "confidence")
pi_band <- predict(model4, newdata = grid, interval = "prediction")

plot(grid$educ, ci_band[, "fit"], type = "l", lwd = 2, col = "blue",
     xlab = "Years of education",
     ylab = "Predicted wage (USD/hour)",
     main = "CI for the mean vs. PI for an individual",
     ylim = range(pi_band))
lines(grid$educ, ci_band[, "lwr"], lty = 2, col = "blue")
lines(grid$educ, ci_band[, "upr"], lty = 2, col = "blue")
lines(grid$educ, pi_band[, "lwr"], lty = 2, col = "red")
lines(grid$educ, pi_band[, "upr"], lty = 2, col = "red")
legend("topleft",
       legend = c("Fitted", "95% CI (mean)", "95% PI (individual)"),
       col = c("blue", "blue", "red"),
       lty = c(1, 2, 2), lwd = c(2, 1, 1))
```

Two features of the plot are worth pointing out:

- The PI (red) is much wider than the CI (blue) at every value of `educ`. The vertical gap between the two bands is, essentially, $\pm t_c \hat\sigma$ --- the contribution of $u_0$.
- Both bands widen at the *extremes* of `educ`, away from the sample mean. Predicting wages for a worker with no schooling, or with 20 years of schooling, is intrinsically less precise than predicting near the sample average. This is the $(x_0 - \bar x)^2 / \mathrm{SST}_x$ term in the variance formula doing its work.

We can now answer the opening question. The narrow interval is the CI for the *mean* wage of workers like the candidate; the wide interval is the PI for *this* candidate's wage. Both are produced by the same model, from the same data, with one keystroke of difference in the call to `predict()`.

## Self-check {.unnumbered}

Six short questions. Try each one before opening the answer.

::: {.callout-tip collapse="true"}
## Q1. Which interval is which?

A recruiter wants to know the *average* wage of all workers in the population who share the candidate's profile. The relevant interval is:

- A. The 95% prediction interval for $y_0$.
- B. The 95% confidence interval for $\mathbb{E}[y \mid x_0]$.
- C. The 95% confidence interval for the slope $\beta_1$.
- D. The OLS residual.

**Answer: B.** The CI for $\mathbb{E}[y \mid x_0]$ targets the conditional *mean* and is the narrower of the two intervals. The PI in option A would be the right choice for "what might *this* candidate actually earn?".
:::

::: {.callout-tip collapse="true"}
## Q2. Why is the PI wider than the CI?

The 95% prediction interval for $y_0$ is wider than the 95% confidence interval for $\mathbb{E}[y \mid x_0]$ because:

- A. The prediction interval uses a larger critical value $t_c$.
- B. The prediction interval includes the variance of the new error $u_0$, the irreducible part of the new draw.
- C. The CI ignores the variance of the OLS estimates entirely.
- D. The PI is built with a smaller sample size.

**Answer: B.** Both intervals use the same $t_c$. The PI adds a "1" inside the bracket in the variance formula, which captures $\operatorname{Var}(u_0) = \sigma^2$ --- something the data can never predict.
:::

::: {.callout-tip collapse="true"}
## Q3. Where are the intervals widest?

Holding the sample fixed, both the CI for the mean and the PI for an individual are *widest* when:

- A. $x_0$ equals the sample mean of the regressor.
- B. The sample size is very large.
- C. $x_0$ is far from the sample mean of the regressor.
- D. The residual variance is zero.

**Answer: C.** Both variances contain the term $(x_0 - \bar x)^2 / \mathrm{SST}_x$, which grows as $x_0$ moves away from $\bar x$. Predictions are most precise near the centre of the data and least precise at the edges; extrapolating beyond the support of the data is dangerous.
:::

::: {.callout-tip collapse="true"}
## Q4. Adjusted $R^2$ and a noise regressor

Adding a pure-noise regressor to an OLS model:

- A. Strictly increases $R^2$ and strictly increases $\bar R^2$.
- B. Cannot decrease $R^2$ but can decrease $\bar R^2$.
- C. Always lowers $R^2$ because the new regressor is uninformative.
- D. Has no effect on either criterion.

**Answer: B.** $R^2$ never falls when a regressor is added. $\bar R^2$ penalises complexity through the degrees-of-freedom correction, so an irrelevant regressor can lower it. AIC and BIC will also worsen.
:::

::: {.callout-tip collapse="true"}
## Q5. Comparing AIC across models

The Akaike Information Criterion can legitimately be used to compare:

- A. Two models estimated on different sub-samples of the data.
- B. A regression of `wage` on `educ` against a regression of `log(wage)` on `educ`.
- C. Two models with the same dependent variable, the same data, and different sets of regressors.
- D. Models estimated by different methods (OLS, IV, GMM) on the same data.

**Answer: C.** AIC (and BIC) are only comparable when the dependent variable, the sample, and the estimation method are the same; otherwise the log-likelihoods are on different scales. The same warning applies to $R^2$ and $\bar R^2$.
:::

::: {.callout-tip collapse="true"}
## Q6. Causality vs. prediction

The model with the smallest AIC is necessarily the best for:

- A. Both prediction *and* causal identification of individual coefficients.
- B. Prediction of $y$ given $x$ --- but not necessarily for identifying the causal effect of a specific regressor.
- C. Causal identification only.
- D. Neither, since AIC is always misleading.

**Answer: B.** AIC is a predictive criterion. Causal identification depends on the assumption $\mathbb{E}[u \mid x] = 0$, which is a statement about the design (random assignment, an instrument, a credible argument), not about model fit. A predictively excellent model can still suffer from omitted-variable bias.
:::

## Exercises {.unnumbered}

**Exercise 5.1 ★ --- The two intervals by hand.** Suppose you have fit $\hat y = 2.0 + 0.7\,x$ on a sample with $n = 100$, $\bar x = 12$, $\mathrm{SST}_x = 200$, and $\hat\sigma^2 = 4$. Take $x_0 = 16$ and use $t_c = 1.98$.

(a) Compute the point prediction $\hat y_0$.
(b) Compute the 95% confidence interval for $\mathbb{E}[y \mid x_0]$.
(c) Compute the 95% prediction interval for $y_0$.
(d) Comment briefly on which interval a *recruiter* would quote and why.

::: {.callout-tip collapse="true"}
## Show answer

(a) $\hat y_0 = 2.0 + 0.7 \times 16 = 13.2$.
(b) $\mathrm{se}(\hat y_0) = 2\sqrt{1/100 + 16/200} = 2\sqrt{0.09} = 0.6$, so CI $= 13.2 \pm 1.98 \times 0.6 = [12.01, 14.39]$.
(c) $\mathrm{se}(f) = 2\sqrt{1 + 0.09} = 2\sqrt{1.09} \approx 2.088$, so PI $= 13.2 \pm 1.98 \times 2.088 = [9.06, 17.34]$.
(d) Both. The CI answers "what is the *average* wage of workers like the candidate?" --- useful for planning aggregate compensation. The PI answers "what might *this* candidate actually earn?" --- useful for an individual job offer. Reporting only one of the two is misleading.
:::

**Exercise 5.2 ★ --- Reading a selection table.** Using `wage1`, estimate the four nested models from the lab and present a 4-by-4 table of $R^2$, $\bar R^2$, AIC, and BIC. Which model wins under each criterion? Do all four criteria agree?

::: {.callout-tip collapse="true"}
## Show answer

In `wage1` the four criteria agree on the ranking: $R^2$ and $\bar R^2$ are highest for `model4`; AIC and BIC are lowest for `model4`. Whenever all four agree the choice is easy. When BIC selects a smaller model than AIC (which can happen with larger samples), the gap reflects BIC's stronger penalty $\ln n > 2$; the choice then depends on whether you prioritise predictive accuracy (AIC) or parsimony (BIC).
:::

**Exercise 5.3 ★ --- A 12-year worker and a 20-year worker.** Using `model4` from the lab, compute the 95% prediction interval for two profiles: (i) `educ = 12, exper = 10, tenure = 5, female = 0, married = 1` and (ii) `educ = 20, exper = 10, tenure = 5, female = 0, married = 1`. Is the second interval wider, narrower, or the same width as the first? Why?

::: {.callout-tip collapse="true"}
## Show answer

The second interval is wider. Twenty years of education is far from the sample mean of `educ` in `wage1` (around 12.5 years), so the $(x_0 - \bar x)' (X'X)^{-1} (x_0 - \bar x)$ term in the variance formula is larger. This is the warning against extrapolation: the further $x_0$ is from the centre of the data, the larger both intervals become, and at some point they become so wide that the prediction is uninformative.
:::

**Exercise 5.4 ★★ --- A garbage variable in your own data.** Take any dataset of your choice (real or simulated). Fit a baseline model with one or two genuine regressors and report $R^2$, $\bar R^2$, AIC, BIC. Add five regressors of pure Gaussian noise (`rnorm(n)`), refit, and report the same four numbers. By how much does each change? Which criterion is fooled, and which is not?

*A full answer is given in the Instructor Edition.*

**Exercise 5.5 ★★ --- Causality vs. prediction with a confound.** Suppose `y` is true grade, `x` is hours studied, and `z` is innate ability. Hours and ability are positively correlated. You fit two models on a large sample: (a) `y ~ x`, (b) `y ~ x + z`. Argue, informally, that:

(i) Model (b) has the higher $R^2$ and higher $\bar R^2$ --- it is the better *predictor*.
(ii) Model (a) gives a *biased* causal estimate of the effect of hours on grades; model (b) does not.
(iii) AIC and BIC will both prefer model (b).

What conclusion do you draw about using AIC/BIC to choose a *causal* model?

*A full answer is given in the Instructor Edition.*

**Exercise 5.6 ★★★ --- Out-of-sample prediction.** Split `wage1` into a training set (the first 400 observations) and a test set (the remaining 126). Fit `model1` through `model4` on the training set. For each model, compute (i) the in-sample $R^2$ on the training set, and (ii) the out-of-sample mean squared error on the test set: $\frac{1}{n_{\text{test}}} \sum_{i \in \text{test}} (y_i - \hat y_i)^2$. Which model wins on in-sample $R^2$? Which wins on out-of-sample MSE? Comment on the relationship between $\bar R^2$/AIC/BIC computed in the training sample and the out-of-sample MSE.

*A full answer is given in the Instructor Edition.*
